Evaluating the Performance and Energy Efficiency of COSMO-ART, a Fully Online Coupled Model System with Numerical Weather Forecast and Chemical Transport Models

### 2014-04-02 paper_enahpc14/Introduction.tex
updates sections 1 & 2; adds file "Conclusion.tex"

COSMO-ART is  computationally much more demanding than  the COSMO core
since a  large number of additional  tracers and processes  have to be
considered.  Thus this model is currently severely-limited in terms of
applicability and expensive in terms of energy consumption.  COSMO has
recently  been  ported  to  GPUs  within the  framework  of  the  High
Performance  and  High  Productivity  Computing (HP2C)  Initiative  to
optimize it  for computational and energy  efficiency.  Although these
developments will  facilitate the  application of COSMO  for numerical
weather prediction and climate  simulations, they do little to address
the  coupling with  the  ART model  extension,  for which  significant
investments  are still required  to take  it to  a similar  level. The
efficiency of ART is being  addressed in the EU Exa2Green project. The
ultimate goal  of the  project is to  deliver a prototype  code, which
provides an energy  efficiency of at least five  times of the baseline
value.   Such   an  implementation   would  allow  the   community  to
investigate critical  questions at  higher resolution and  over longer
periods, at reduced cost  to the environment.  In Sec.~\ref{sec:1}, we
introduce the regional  atmospheric model COSMO-ART \citep{Vogel-2009}
that  accounts for feedbacks  between chemistry,  aerosols, radiation,
and   clouds.   Sec.~\ref{sec:2}  and   Sec.~\ref{sec:3}  successively
describe the  hardware and measurement  devices used in  evaluation as
well  as  energy-aware techniques  employed  to  determine the  energy
footprint  of  the  considered  application.   Sec.~\ref{sec:4}  shows
benchmark  results from  the  Exa2Green project  and highlights  areas
where improvements  will be  necessary for the  baseline optimisation.
Finally, we  conclude and give directions for  further developments of
the modeling system in Sec.~\ref{concl}.

>>>>

COSMO-ART is  computationally much more demanding than  the COSMO core
since a  large number of additional  tracers and processes  have to be
considered.  Thus this model is currently severely-limited in terms of
applicability and expensive in terms of energy consumption.  COSMO has
recently  been  ported  to  GPUs  within the  framework  of  the  High
Performance  and  High  Productivity  Computing (HP2C)  Initiative  to
optimize it  for computational and energy  efficiency.  Although these
developments will  facilitate the  application of COSMO  for numerical
weather prediction and climate  simulations, they do little to address
the  coupling with  the  ART model  extension,  for which  significant
investments  are still required  to take  it to  a similar  level. The
efficiency of ART is being  addressed in the EU Exa2Green project. The
ultimate goal  of the  project is to  deliver a prototype  code, which
provides an energy  efficiency of at least five  times of the baseline
value.   Such   an  implementation   would  allow  the   community  to
investigate critical  questions at  higher resolution and  over longer
periods, at reduced cost  to the environment.  In Sec.~\ref{sec:1}, we
introduce the regional  atmospheric model COSMO-ART \citep{Vogel-2009}
that  accounts for feedbacks  between chemistry,  aerosols, radiation,
and clouds and additionally specify its technical setup related to the
investigated    performance    and     energy    evaluation    methods.
Sec.~\ref{sec:2}  describes the different  HPC systems  and associated
power  measurement  equipments used  for  conducting the  benchmarking
study and  presents energy-aware techniques employed  to determine the
energy  footprint  of  the considered  application.   Sec.~\ref{sec:3}
gathers benchmark  results from  the Exa2Green project  and highlights
areas where improvements will be necessary for the subsequent baseline
refactoring.  Some related work  will be discussed in Sec.~\ref{sec:4}
Finally, we  conclude with practical guidelines  to identifying energy
saving opportunities  and give directions for  further developments of
the modeling system in Sec.~\ref{concl}.

### 2014-04-02 paper_enahpc14/Sec1:COSMO-ART.tex
updates sections 1 & 2; adds file "Conclusion.tex"

\subsection{Model description}
\label{subsec:1.1}

COSMO-ART  (ART   stands  for  Aerosols  and   Reactive  Trace  gases,
\citealp{Vogel-2009}),  developed  at KIT  Karlsruhe  (Germany), is  a
regional  to  continental scale  model  coupled  online  to the  COSMO
numerical    weather    prediction    (NWP)    and    climate    model
\citep{Baldauf-2011}. COSMO is used  by several European countries for
operational weather  forecast. The  gaseous chemistry in  COSMO-ART is
solved by  a modified version  of the Regional Acid  Deposition Model,
Version  2 (RADM2)  mechanism \citep{Stockwell-1990},  which  has been
extended   to    describe   secondary   organic    aerosol   formation
\citep{Schell-2001, Athanasopoulou-2013} and hydroxyl recycling due to
isoprene chemistry \citep{Geiger-2003}  and heterogeneous reactions as
hydrolysis  of N2O5.  Aerosols are  represented by  the  modal aerosol
module MADE \citep{Ackermann-1998},  improved by explicit treatment of
soot aging  through condensation of  inorganic \citep{Riemer-2003} and
organic substances (MADEsoot).

The  five modes that  represent the  aerosol population  contain: pure
soot, secondary mixtures of sulfates, nitrates, ammonium, organics and
water (nucleation  and accumulation) and the internal  mixtures of all
these species in  both modes. Separate fine and  coarse emission modes
for  sea-salt,  dust  \citep{Stanelle-2010},  and  rest  anthropogenic
species  are treated by  six additional  modes.  Specific  modules are
included    to   simulate    the   dispersion    of    pollen   grains
\citep{Vogel-2008}   and  other  biological   particles.   Meteorology
affected emissions are also online coupled within the model system.

\subsection{Model setup}
\label{subsec:1.2}

>>>>

\subsection{Model description}
\label{subsec:1.1}
COSMO-ART  (ART  stands  for   Aerosols  and  Reactive  Trace  gases),
developed  at  KIT  Karlsruhe  \citep{Vogel-2009}, is  a  regional  to
continental scale model coupled  online to the COSMO numerical weather
prediction  (NWP) and  climate  model \citep{Baldauf-2011}.   Physical
processes  like  transport,  turbulent  diffusion,  and  dry  and  wet
deposition  are  treated  together  with  photochemistry  and  aerosol
dynamics  using the  modal approach.

Aerosols dynamics  are simulated  with the modal  aerosol microphysics
module  MADE \citep{Ackermann-1998}, expanded  in MADEsoot  to feature
explicit  treatment of  soot aging  through condensation  of inorganic
\citep{Riemer-2003} and organic substances. MADE describes the aerosol
population  through  five   modes  representing  sub-micron  particles
consisting of sulphate, ammonium, nitrate, particulate organic matter,
water and soot  \citep{Riemer-2004} in a range of  mixing state. These
modes are coupled  with the gas phase by  condensation and nucleation,
and are  strongly influenced by  anthropogenic emissions of  gases and
particles. MADEsoot  describes the sub-micrometer  aerosol population,
composed  of sea  salt \citep{Lundgren-2013},  mineral  dust particles
\citep{Vogel-2006,  Stanelle-2010},   by  means  of   six  interacting
log-normal modes.  For each  mode, mass contributions and total number
concentration are prognostic  quantities, while the standard deviation
is fixed.

Specific  modules are included  to simulate  the dispersion  of pollen
grains    \citep{Vogel-2008}   and    other    biological   particles.
Meteorologically-influenced emissions  are also online  coupled within
the  model  system.  The  biogenic  VOC  (volatile organic  compounds)
emissions are  calculated as functions of  the land use  type based on
the Global  Land Cover 2000  dataset and the modeled  temperatures and
radiative fluxes \citep{Vogel-1995}.

The gaseous chemistry in COSMO-ART  is solved by a modified version of
the  Regional  Acid  Deposition  Model, Version  2  (RADM2)  mechanism
\citep{Stockwell-1990},  which  is   extended  to  describe  secondary
organic aerosol formation based on a volatile basis set (VBS) approach
\citep{Athanasopoulou-2013}  and  hydroxyl  radical recycling  due  to
isoprene chemistry \citep{Geiger-2003}.  The thermodynamic equilibrium
between the  gas and particulate  phases of the inorganic  material is
achieved through the  ISORROPIA II module \citep{Fountoukis-2007}.

COSMO-ART  is  fully  online-coupled,  and  allows  for  feedbacks  of
aerosols  on  temperature,  radiation  and cloud  condensation  nuclei
(CCN).   Analytical  description  of  modules  incorporated  for  this
purpose  exists in  \citep{Vogel-2009,  Bangert-2011}.  The  radiation
scheme used  within the  model to calculate  the vertical  profiles of
shortwave and longwave radiative fluxes is GRAALS \citep{Ritter-1992}.
In order to account for  the interaction of aerosol particles with the
cloud microphysics and radiation,  COSMO-ART uses the two-moment cloud
microphysics  scheme of  Seifert and  Beheng  \citep{Seifert-2006} and
comprehensive parameterizations for cloud condensation and ice crystal
nucleation  \citep{Bangert-2011, Bangert-2012}.   The system  of stiff
ordinary differential equations described by chemical reactions in the
aqueous-phase together with the transfer reactions is solved using the
Kinetic PreProcessor (KPP, \citealp{Damian-2002}).

Detailed  model  description  can   be  found  in  the  aforementioned
publications  as   well  as  in   \citep{Stanelle-2010,  Bangert-2012,
  Knote-2011, Knote-2013}.

\subsection{Model setup}
\label{subsec:1.2}
Establishing effective energy performance benchmarking of a code under
intense development such as COSMO-ART is a challenging task because of
the  absolute necessity that  results must  be reproducible  within an
expected  variance for the  duration of  the Exa2Green  project. After
consultation  with  the  climate  community  to  properly  define  the
baseline,  it was  necessary to  find a  run-configuration  capable of
being recreated in all subsequent versions of the code.

Thus, we give here a brief  overview of the main features of the model
setup for  our benchmark's daily  runs.  Three-dimensional simulations
are  performed over  Europe  for April  13th,  which is  close to  the
equinox and thus  brings benefits of having approximately  half day of
night  and half  day  of  sun exposure,  therefore  ensuring a  proper
activation of the chemistry cycle. We consider daily 24-hour forecasts
without  any  previous spin-up  simulations.   The calculation  domain
corresponds to  the CORDEX-EU-44  domain and is  covered by a  grid of
$222\times   216$    points   with   a    horizontal   resolution   of
$0.22\,^{\circ}$,  i.e.  50  km in  both directions,  and  40 vertical
layers.   Boundary  and initial  conditions,  emmissions and  external
parameters  were obtained by  the INT2LM  pre-processing interpolation
tool, which  performed the usual interpolation from  coarse grid model
data, and  INT2LM\_trcr\_ART, which consists mostly  of the definition
of a large number of  tracers.  The meteorological initial and bounday
conditions are  obtained by  the the ECMWF  global spectral  model IFS
with an update  frequency of 3h.  Boundary data  for gas-phase species
are taken from IFS-MOZART  output at 6h temporal resolution.  Multiple
runs of COSMO-ART were  performed to illustrate the reproducibility of
the baseline, and quantify  the significant uncertainties in the power
measurement,   as   dictated   by   the  available   technology   (see
subsection~\ref{subsec:3.2}).   The run-configuration  incorporates 34
2-d and 45 3-d fields to be written out every hour and doesn't include
data assimilation methods.

This COSMO-ART  version is configured  to deal with a  semi Lagrangian
horizontal advection scheme  with tricubic interpolation and selective
filling diffusion option in combination with the Runge-Kutta dynamical
core. It  also makes  use of the  Kinetic PreProcessor solver  for the
resolution of  atmospheric chemistry ordinary  differential equations.
Concerning the  modelling of wet deposition in  aerosols, the baseline
has  only  indirect  cloud  feedbacks  but  doesn't  include  in-cloud
scavenging  (rainout)   and  below-cloud  scavenging   (washout)  yet.
Amongst   physical  parameterizations,   precipitation   formation  is
performed  by a  two-moment  cloud microphysics  scheme  instead of  a
classical  bulk microphysics  scheme.

### 2014-04-02 paper_enahpc14/paper.tex
updates sections 1 & 2; adds file "Conclusion.tex"

\begin{abstract}
  In this paper we present  COSMO-ART, an extension of the operational
  weather  forecast  model  of   the  German  Weather  Service  (DWD),
  developed for  the evaluation of the interactions  of reactive gases
  and aerosol particles  with the state of atmosphere  at the regional
  scale. It  includes secondary aerosols,  directly emitted components
  like  soot,  mineral  dust,  sea  salt and  biological  material  as
  pollen. Processes such  as emissions, coagulation, condensation, dry
  deposition,  wet removal,  and sedimentation  of aerosols  are taken
  into account.   The overall performance  of this application  on HPC
  systems is  analysed by a  profiling and tracing study  to determine
  hotspots  and  identify   critical  paths.   Moreover,  we  describe
  measurement devices and energy-aware techniques employed to evaluate
  the  energy  footprint of  the  considered  application  and to  get
  detailed  insights about  power bottlenecks.   Our motivation  is to
  improve  corresponding  code sections  to  sustain high  performance
  while minimizing energy-to-solution.  This preliminary work sets the
  basis for subsequent studies  to tackle challenges related to energy
  efficient  high  performance  computing  in  the  framework  of  the
  Exa2Green project (\url{http://exa2green.eu/}).

@@

\section{Hardware description}
\label{sec:2}

\section{Power measurement systems}
\label{sec:3}
\input{Sec3:Power-measurement.tex}

\section{Benchmark results}
\label{sec:4}

\section{Conclusion}
\label{concl}

>>>>

\begin{abstract}
  In this paper we present  COSMO-ART, an extension of the operational
  weather  forecast  model  of   the  German  weather  service  (DWD),
  developed for  the evaluation of the interactions  of reactive gases
  and aerosol particles  with the state of atmosphere  at the regional
  scale. It  includes secondary aerosols,  directly emitted components
  like  soot,  mineral  dust,  sea  salt and  biological  material  as
  pollen. Processes such  as emissions, coagulation, condensation, dry
  deposition,  wet removal,  and sedimentation  of aerosols  are taken
  into account.   The overall performance  of this application  on HPC
  systems is analysed  by a profiling study to  determine hotspots and
  identify critical paths.   Moreover, we describe measurement devices
  and  energy-aware   techniques  employed  to   evaluate  the  energy
  footprint of the considered application and to get detailed insights
  about power bottlenecks.  Our motivation is to improve corresponding
  code   sections  to  sustain   high  performance   while  minimizing
  energy-to-solution.   This  preliminary  work  sets  the  basis  for
  subsequent studies to tackle  challenges related to energy efficient
  high performance computing in the framework of the Exa2Green project
  (\url{http://exa2green.eu/}).

@@

\section{Power-performance measurement framework}
\label{sec:2}
\input{Sec2:Power-measurement.tex}

\section{Evaluation}
\label{sec:3}
\input{Sec3:Evaluation.tex}

\section{Related work}
\label{sec:4}
\input{Sec4:Related.tex}

\section{Conclusion}
\label{concl}
\input{Conclusion.tex}
